STJ2012 analysis

### 2016-08-11 analysis/STJ2012_analysis.Rmd
make networks based on bysample OTUs

# Data analysis
## Community composition barplots
Here the composition of each sample is plotted as a horizontal bar, sorted by species and shore. Each segment of the bar represents an individual OTU. In the lefthand plot, all OTUs are colored by clade, in the right OTU, all OTUs are given a unique color.
```{r cladecomp97, cache=T, fig.height=6, fig.width=8}
# Define function to plot Symbiodinium community composition
composition <- function(phy, col, legend=T) {
  samdat <- data.frame(sample_data(phy))
  samdat$Genus <- factor(samdat$Genus, levels=rev(levels(samdat$Genus)))
  samdat$Species <- factor(samdat$Species, levels=rev(levels(samdat$Species)))
  samdat$Shore <- factor(samdat$Shore, levels=rev(levels(samdat$Shore)))
  samdat <- samdat[with(samdat, order(Genus, Species, Shore)), ]
  typerelabund <- as.matrix(otu_table(phy)[order(data.frame(tax_table(phy))$Subtype), 
                                           rownames(samdat)])
  shorebreaks <- c(as.character(samdat$Shore), "X")==c("X", as.character(samdat$Shore))
  shorebreaks <- which(shorebreaks==F) - 1
  spbreaks <- c(which(duplicated(samdat$Species)==F) - 1, nrow(samdat))
  # Make Barplot
  barplot(typerelabund, horiz=T, space=0, axes=F,axisnames=F, yaxs="i", col=col)
  rect(0, 0, par("usr")[2], par("usr")[4], lwd=1, xpd=T)
  axis(side=1, at=seq(0, 1, 0.1), line=0, tck=-0.025, mgp=c(0,0.25,0), cex.axis=0.7)
  mtext(side=1, "Relative abundance", cex=0.7, line=1)
  # Add legend
  if (legend==T) {
    legend(x=par("usr")[2]/2, y=par("usr")[4], xjust=0.5, yjust=0.25, horiz=T, bty="n", xpd=T, 
           cex=0.7, legend=c("A", "B", "C", "D", "F"), fill=taxcolors, x.intersp=0.5)
    legend(x=par("usr")[2]-0.03, y=par("usr")[4], xjust=0, yjust=0.1, bty="n", xpd=T, cex=0.6, 
           pt.cex=0.6, legend=c("N Shore", "S Shore"), fill=c(0, "gray40"), y.intersp=0.7, 
           x.intersp=0.3)
  }
  # Add grouping bars for Shore
  for (i in 1:length(shorebreaks)) {
    lines(c(0, 1), c(shorebreaks[i], shorebreaks[i]), lty=2, lwd=0.25)
    rect(1.01, shorebreaks[i], 1.04, shorebreaks[i+1], col=rep(c("gray40", 0), 50)[i], 
         lwd=0.25, xpd=T)
  }
  # Add lines to separate species and species names
  for (i in 1:length(spbreaks)) {
    lines(c(0, 1.07), c(spbreaks[i], spbreaks[i]), xpd=T, type="l", lwd=0.4)
    text(1.03, (spbreaks[i] + spbreaks[i+1]) / 2, xpd=T, pos=4, cex=0.6,
        labels=paste(samdat$Genus[which(duplicated(samdat$Species)==F)][i], "\n",
                    samdat$Species[which(duplicated(samdat$Species)==F)][i], sep=""))
  }
}

@@

# Plot composition of 97% within-sample OTUs colored by clade
composition(phy97bs.f.p, col=taxcolors[data.frame(tax_table(phy97bs.f.p))[order(data.frame(tax_table(phy97bs.f.p))$Subtype), ]$Clade], legend=T)

@@

# Network analysis from tutorial
```{r networks}
make.net <- function(physeq, n) {
  # Remove taxa that are not present in subset
  physeq <- prune_taxa(rowSums(otu_table(physeq))!=0, physeq)
  # Convert OTU table to "edges" table (weight = relative abundance)
  otudf <- data.frame(t(otu_table(physeq)), check.names=F)
  edges <- setNames(melt(as.matrix(otudf)), nm=c("source", "target", "weight"))  # Melt data
  edges <- edges[edges$weight>n, ]
  # Create "nodes" table from sample and tax data
  sdf <- data.frame(sample_data(physeq))
  sdf$id <- rownames(sdf)
  tdf <- data.frame(tax_table(physeq))
  tdf$id <- rownames(tdf)
  nodes <- merge(sdf, tdf, all=T)
  # Create network
  net <- graph_from_data_frame(d=edges, vertices=nodes, directed=F)
  return(net)
}
dstr.net <- make.net(dstr97.f.p, 0)

plot(dstr.net,
     edge.curved=0.1, 
     edge.width=10*(E(dstr.net)$weight)^0.2,
     vertex.label=substr(V(dstr.net)$Subtype, 1, 4),
     vertex.color=taxcolors[factor(V(dstr.net)$Clade, levels=c("A","B","C","D","F","G"))])

pfur.net <- make.net(pfur97.f.p, 0)
plot(pfur.net,
     edge.curved=0.1, 
     edge.width=10*(E(pfur.net)$weight)^0.2,
     vertex.label=substr(V(pfur.net)$Subtype, 1, 4),
     vertex.color=taxcolors[factor(V(pfur.net)$Clade, levels=c("A","B","C","D","F","G"))])

oann97.f.p <- prune_taxa(rowSums(otu_table(oann97.f.p))!=0, oann97.f.p)
oann.net <- make.net(oann97.f.p, 0)
plot(oann.net,
     edge.curved=0.1, 
     edge.width=10*(E(oann.net)$weight)^0.2,
     vertex.label=substr(V(oann.net)$Subtype, 1, 4),
     vertex.color=taxcolors[factor(V(oann.net)$Clade, levels=c("A","B","C","D","F","G"))])

malc.net <- make.net(malc97.f.p, 0)
plot(malc.net,
     edge.curved=0.1, 
     edge.width=10*(E(malc.net)$weight)^0.2,
     vertex.label=substr(V(malc.net)$Subtype, 1, 4),
     vertex.color=taxcolors[factor(V(malc.net)$Clade, levels=c("A","B","C","D","F","G"))])

ssid.net <- make.net(ssid97.f.p, 0)
plot(ssid.net,
     edge.curved=0.1, 
     edge.width=10*(E(ssid.net)$weight)^0.2,
     vertex.label=substr(V(ssid.net)$Subtype, 1, 4),
     vertex.color=taxcolors[factor(V(ssid.net)$Clade, levels=c("A","B","C","D","F","G"))])

ffra.net <- make.net(ffra97.f.p, 0)
plot(ffra.net,
     edge.curved=0.1, 
     edge.width=10*(E(ffra.net)$weight)^0.2,
     vertex.label=substr(V(ffra.net)$Subtype, 1, 4),
     vertex.color=taxcolors[factor(V(ffra.net)$Clade, levels=c("A","B","C","D","F","G"))])

srad.net <- make.net(srad97.f.p, 0)
plot(srad.net,
     edge.curved=0.1, 
     edge.width=10*(E(srad.net)$weight)^0.2,
     vertex.label=substr(V(srad.net)$Subtype, 1, 4),
     vertex.color=taxcolors[factor(V(srad.net)$Clade, levels=c("A","B","C","D","F","G"))])

past.net <- make.net(past97.f.p, 0)
plot(past.net,
     edge.curved=0.1, 
     edge.width=10*(E(past.net)$weight)^0.2,
     vertex.label=substr(V(past.net)$Subtype, 1, 4),
     vertex.color=taxcolors[factor(V(past.net)$Clade, levels=c("A","B","C","D","F","G"))])

dcyl.net <- make.net(dcyl97.f.p, 0)
plot(dcyl.net,
     edge.curved=0.1, 
     edge.width=10*(E(dcyl.net)$weight)^0.2,
     vertex.label=substr(V(dcyl.net)$Shore, 1, 4),
     vertex.color=taxcolors[factor(V(dcyl.net)$Clade, levels=c("A","B","C","D","F","G"))])

mcav.net <- make.net(mcav97.f.p, 0)
plot(mcav.net,
     edge.curved=0.1, 
     edge.width=10*(E(mcav.net)$weight)^0.2,
     vertex.label=substr(V(mcav.net)$Subtype, 1, 4),
     vertex.color=taxcolors[factor(V(mcav.net)$Clade, levels=c("A","B","C","D","F","G"))])

@@

# Get network with one node for each species.
# Edge weight is the mean relative abundance of OTU.
# Convert OTU table to "edges" table (weight = relative abundance)
  otudf <- data.frame(t(otu_table(phy97.f.p)), check.names=F)
  # Average relative abundance of OTU within species
  #sp.ag <- aggregate(otudf, by=list(Species=data.frame(sample_data(phy97.f.p))$Species), FUN=mean)
  # Percent of samples of species that contain OTU at more than 1% relative abundance
  sp.ag <- aggregate(otudf, by=list(Species=data.frame(sample_data(phy97.f.p))$Species), 
                     FUN=function(x) length(x[x>0.01])/length(x))

  rownames(sp.ag) <- sp.ag$Species
  sp.ag <- sp.ag[, -1]
  edges <- setNames(melt(as.matrix(sp.ag)), nm=c("source", "target", "weight"))  # Melt data
  edges <- droplevels(edges[edges$weight>0, ])
  # Create "nodes" table from sample and tax data
  sdf <- data.frame(id=levels(data.frame(sample_data(phy97.f.p))$Species))
  tdf <- data.frame(tax_table(phy97.f.p))[rownames(data.frame(tax_table(phy97.f.p))) %in% edges$target, ]
  tdf$id <- rownames(tdf)
  nodes <- merge(sdf, tdf, all=T)
  nodes$type <- ifelse(is.na(nodes$Clade), 1, 2)
  # Create network
  net <- graph_from_data_frame(d=edges, vertices=nodes, directed=F)

  # Best one for now ---------
  V(net)$shape <- c("square", "circle")[factor(V(net)$type)]
  V(net)$size <- ifelse(is.na(V(net)$Clade), 15, 10*sqrt(degree(net))) #5*sqrt(degree(net))
  V(net)$label <- ifelse(is.na(V(net)$Clade), names(V(net)),  
                         unlist(lapply(strsplit(V(net)$Subtype, split="_"), "[", 1)))
  V(net)$label <- str_replace_all(V(net)$label, c("alcicornis"="M.alc.", "annularis"="O.ann",
                                                  "astreoides"="P.ast.", "cavernosa"="M.cav.",
                                                  "cylindrus"="D.cyl.", "fragum"="F.fra.", 
                                                  "furcata"="P.fur.", "radians"="S.rad.", 
                                                  "siderea"="S.sid.", "strigosa"="D.str."))
  V(net)$color <- ifelse(is.na(V(net)$Clade), "white",
                         taxcolors[factor(V(net)$Clade, levels=c("A","B","C","D","F","G"))])
  E(net)$color <- "gray60"
  E(net)$width <- 15 * E(net)$weight
  set.seed(12347)
  l <- layout.fruchterman.reingold(net)
  l <- norm_coords(l, ymin=-1, ymax=1, xmin=-1, xmax=1)
  plot(net, rescale=F, layout=l*1, edge.curved=0.1, vertex.label.cex=V(net)$size/20,
       vertex.label.color="black")
  legend(x=-1.05, y=-0.92, c("CladeA", "CladeB", "CladeC", "CladeD", "CladeF", "CladeG"), pch=21,
       col="#777777", pt.bg=taxcolors, pt.cex=1.1, cex=.6, bty="n", ncol=2, text.width=0.2)
  # --------------------------

>>>>

# Data analysis
## Community composition barplots
Here the composition of each sample is plotted as a horizontal bar, sorted by species and shore. Each segment of the bar represents an individual OTU. In the lefthand plot, all OTUs are colored by clade, in the right OTU, all OTUs are given a unique color.
```{r cladecomp97, cache=T, fig.height=6, fig.width=8}
# Define function to plot Symbiodinium community composition
composition <- function(phy, col, legend=T) {
  samdat <- data.frame(sample_data(phy))
  samdat$Genus <- factor(samdat$Genus, levels=rev(levels(samdat$Genus)))
  samdat$Species <- factor(samdat$Species, levels=rev(levels(samdat$Species)))
  samdat$Shore <- factor(samdat$Shore, levels=rev(levels(samdat$Shore)))
  samdat <- samdat[with(samdat, order(Genus, Species, Shore)), ]
  typerelabund <- as.matrix(otu_table(phy)[order(data.frame(tax_table(phy))$Subtype), 
                                           rownames(samdat)])
  shorebreaks <- c(as.character(samdat$Shore), "X")==c("X", as.character(samdat$Shore))
  shorebreaks <- which(shorebreaks==F) - 1
  spbreaks <- c(which(duplicated(samdat$Species)==F) - 1, nrow(samdat))
  # Make Barplot
  barplot(typerelabund, horiz=T, space=0, axes=F,axisnames=F, yaxs="i", col=col)
  rect(0, 0, par("usr")[2], par("usr")[4], lwd=1, xpd=T)
  axis(side=1, at=seq(0, 1, 0.1), line=0, tck=-0.025, mgp=c(0,0.25,0), cex.axis=0.7)
  mtext(side=1, "Relative abundance", cex=0.7, line=1)
  # Add legend
  if (legend==T) {
    legend(x=par("usr")[2]/2, y=par("usr")[4], xjust=0.5, yjust=0.25, horiz=T, bty="n", xpd=T, 
           cex=0.7, legend=c("A", "B", "C", "D", "F", "G"), fill=taxcolors, x.intersp=0.5)
    legend(x=par("usr")[2]-0.03, y=par("usr")[4], xjust=0, yjust=0.1, bty="n", xpd=T, cex=0.6, 
           pt.cex=0.6, legend=c("N Shore", "S Shore"), fill=c(0, "gray40"), y.intersp=0.7, 
           x.intersp=0.3)
  }
  # Add grouping bars for Shore
  for (i in 1:length(shorebreaks)) {
    lines(c(0, 1), c(shorebreaks[i], shorebreaks[i]), lty=2, lwd=0.25)
    rect(1.01, shorebreaks[i], 1.04, shorebreaks[i+1], col=rep(c("gray40", 0), 50)[i], 
         lwd=0.25, xpd=T)
  }
  # Add lines to separate species and species names
  for (i in 1:length(spbreaks)) {
    lines(c(0, 1.07), c(spbreaks[i], spbreaks[i]), xpd=T, type="l", lwd=0.4)
    text(1.03, (spbreaks[i] + spbreaks[i+1]) / 2, xpd=T, pos=4, cex=0.6,
        labels=paste(samdat$Genus[which(duplicated(samdat$Species)==F)][i], "\n",
                    samdat$Species[which(duplicated(samdat$Species)==F)][i], sep=""))
  }
}

@@

# Plot composition of 97% within-sample OTUs colored by clade
composition(phy97bs.f.p, col=taxcolors[factor(data.frame(tax_table(phy97bs.f.p))[order(data.frame(tax_table(phy97bs.f.p))$Subtype), ]$Clade, levels=c("A","B","C","D","F","G"))], legend=T)

@@

# Network analysis from tutorial
```{r networks}
make.net <- function(physeq, n) {
  # Remove taxa that are not present in subset
  physeq <- prune_taxa(rowSums(otu_table(physeq))!=0, physeq)
  # Convert OTU table to "edges" table (weight = relative abundance)
  otudf <- data.frame(t(otu_table(physeq)), check.names=F)
  edges <- setNames(melt(as.matrix(otudf)), nm=c("source", "target", "weight"))  # Melt data
  edges <- edges[edges$weight>n, ]
  # Create "nodes" table from sample and tax data
  sdf <- data.frame(sample_data(physeq))
  sdf$id <- rownames(sdf)
  tdf <- data.frame(tax_table(physeq))
  tdf$id <- rownames(tdf)
  nodes <- merge(sdf, tdf, all=T)
  # Create network
  net <- graph_from_data_frame(d=edges, vertices=nodes, directed=F)
  return(net)
}
dstr.net <- make.net(dstr97bs.f.p, 0)

set.seed(5453890)
plot(dstr.net,
     edge.curved=0.1, 
     edge.width=10*(E(dstr.net)$weight)^0.2,
     vertex.label=substr(V(dstr.net)$Subtype, 1, 4),
     vertex.color=taxcolors[factor(V(dstr.net)$Clade, levels=c("A","B","C","D","F","G"))])

pfur.net <- make.net(pfur97bs.f.p, 0)
plot(pfur.net,
     edge.curved=0.1, 
     edge.width=10*(E(pfur.net)$weight)^0.2,
     vertex.label=substr(V(pfur.net)$Subtype, 1, 4),
     vertex.color=taxcolors[factor(V(pfur.net)$Clade, levels=c("A","B","C","D","F","G"))])

oann.net <- make.net(oann97bs.f.p, 0)
plot(oann.net,
     edge.curved=0.1, 
     edge.width=10*(E(oann.net)$weight)^0.2,
     vertex.label=substr(V(oann.net)$Subtype, 1, 4),
     vertex.color=taxcolors[factor(V(oann.net)$Clade, levels=c("A","B","C","D","F","G"))])

malc.net <- make.net(malc97bs.f.p, 0)
plot(malc.net,
     edge.curved=0.1, 
     edge.width=10*(E(malc.net)$weight)^0.2,
     vertex.label=substr(V(malc.net)$Subtype, 1, 4),
     vertex.color=taxcolors[factor(V(malc.net)$Clade, levels=c("A","B","C","D","F","G"))])

set.seed(9147)
ssid.net <- make.net(ssid97bs.f.p, 0)
plot(ssid.net,
     edge.curved=0.1, 
     edge.width=10*(E(ssid.net)$weight)^0.2,
     vertex.label=substr(V(ssid.net)$Subtype, 1, 4),
     vertex.color=taxcolors[factor(V(ssid.net)$Clade, levels=c("A","B","C","D","F","G"))])

ffra.net <- make.net(ffra97bs.f.p, 0)
plot(ffra.net,
     edge.curved=0.1, 
     edge.width=10*(E(ffra.net)$weight)^0.2,
     vertex.label=substr(V(ffra.net)$Subtype, 1, 4),
     vertex.color=taxcolors[factor(V(ffra.net)$Clade, levels=c("A","B","C","D","F","G"))])

srad.net <- make.net(srad97bs.f.p, 0)
plot(srad.net,
     edge.curved=0.1, 
     edge.width=10*(E(srad.net)$weight)^0.2,
     vertex.label=substr(V(srad.net)$Subtype, 1, 4),
     vertex.color=taxcolors[factor(V(srad.net)$Clade, levels=c("A","B","C","D","F","G"))])

past.net <- make.net(past97bs.f.p, 0)
plot(past.net,
     edge.curved=0.1, 
     edge.width=10*(E(past.net)$weight)^0.2,
     vertex.label=substr(V(past.net)$Subtype, 1, 4),
     vertex.color=taxcolors[factor(V(past.net)$Clade, levels=c("A","B","C","D","F","G"))])

dcyl.net <- make.net(dcyl97bs.f.p, 0)
plot(dcyl.net,
     edge.curved=0.1, 
     edge.width=10*(E(dcyl.net)$weight)^0.2,
     vertex.label=substr(V(dcyl.net)$Subtype, 1, 4),
     vertex.color=taxcolors[factor(V(dcyl.net)$Clade, levels=c("A","B","C","D","F","G"))])

mcav.net <- make.net(mcav97bs.f.p, 0)
plot(mcav.net,
     edge.curved=0.1, 
     edge.width=10*(E(mcav.net)$weight)^0.2,
     vertex.label=substr(V(mcav.net)$Subtype, 1, 4),
     vertex.color=taxcolors[factor(V(mcav.net)$Clade, levels=c("A","B","C","D","F","G"))])

@@

# Get network with one node for each species.
# Edge weight is the mean relative abundance of OTU.
# Convert OTU table to "edges" table (weight = relative abundance)
  otudf <- data.frame(t(otu_table(phy97bs.f.p)), check.names=F)
  # Average relative abundance of OTU within species
  #sp.ag <- aggregate(otudf, by=list(Species=data.frame(sample_data(phy97.f.p))$Species), FUN=mean)
  # Percent of samples of species that contain OTU at more than 1% relative abundance
  sp.ag <- aggregate(otudf, by=list(Species=data.frame(sample_data(phy97bs.f.p))$Species), 
                     FUN=function(x) length(x[x>0.01])/length(x))

  rownames(sp.ag) <- sp.ag$Species
  sp.ag <- sp.ag[, -1]
  edges <- setNames(melt(as.matrix(sp.ag)), nm=c("source", "target", "weight"))  # Melt data
  edges <- droplevels(edges[edges$weight>0, ])
  # Create "nodes" table from sample and tax data
  sdf <- data.frame(id=levels(data.frame(sample_data(phy97bs.f.p))$Species))
  tdf <- data.frame(tax_table(phy97bs.f.p))[rownames(data.frame(tax_table(phy97bs.f.p))) %in% edges$target, ]
  tdf$id <- rownames(tdf)
  nodes <- merge(sdf, tdf, all=T)
  nodes$type <- ifelse(is.na(nodes$Clade), 1, 2)
  # Create network
  net <- graph_from_data_frame(d=edges, vertices=nodes, directed=F)

  # Best one for now ---------
  V(net)$shape <- c("square", "circle")[factor(V(net)$type)]
  V(net)$size <- ifelse(is.na(V(net)$Clade), 15, 10*sqrt(degree(net))) #5*sqrt(degree(net))
  V(net)$label <- ifelse(is.na(V(net)$Clade), names(V(net)),  
                         unlist(lapply(strsplit(V(net)$Subtype, split="_"), "[", 1)))
  V(net)$label <- str_replace_all(V(net)$label, c("alcicornis"="M.alc.", "annularis"="O.ann",
                                                  "astreoides"="P.ast.", "cavernosa"="M.cav.",
                                                  "cylindrus"="D.cyl.", "fragum"="F.fra.", 
                                                  "furcata"="P.fur.", "radians"="S.rad.", 
                                                  "siderea"="S.sid.", "strigosa"="D.str."))
  V(net)$color <- ifelse(is.na(V(net)$Clade), "white",
                         taxcolors[factor(V(net)$Clade, levels=c("A","B","C","D","F","G"))])
  E(net)$color <- "gray60"
  E(net)$width <- 15 * E(net)$weight
  set.seed(12347)
  set.seed(12364)
  set.seed(12374)
  l <- layout.fruchterman.reingold(net)
  l <- norm_coords(l, ymin=-1, ymax=1, xmin=-1, xmax=1)
  par(mar=c(0,0,0,0))
  plot(net, rescale=F, layout=l*1, edge.curved=0.1, vertex.label.cex=V(net)$size/20,
       vertex.label.color="black")
  #legend(x=-1.05, y=1.1, c("CladeA", "CladeB", "CladeC", "CladeD", "CladeF", "CladeG"), pch=21,
  #     col="#777777", pt.bg=taxcolors, pt.cex=1.1, cex=.6, bty="n", ncol=2, text.width=0.2)
```
